---
always_allow_html: true
output: github_document

---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(tidyverse)
library(knitr)
library(kableExtra)
```
# crmReporter

The `crmReporter` package is designed to provide a simple interface between `crmPack` (and other MCMC packages) and [the tidyverse](www.tidyverse.org) and to augment the reporting facilities provided by `crmPack`, `coda`, `runjags` and other packages relating to MCMC simulation.


## Installation
`crmReporter` is available from GitHub.  To install it, use
```{r, eval=FALSE}
devtools::install_github("PuzzledFace/crmReporter")
```

## Creating tidy data
Creating tidy data with `crmReporter` is simple: use the `tidy()` verb.  Tidy works with `mcmc`s, `mcmc.list`s, `runjags`s and `runjags.list`s.
```{r}
library(crmReporter)

data("oCRMPosteriorShort")
x <- oCRMPosteriorShort %>% tidy()
x %>% head()
```

You can use the output from `tidy()` as input to other `crmReporter` functions or to your own code built within the Tidyverse.  Here are some sample outputs from `crmReporter` functions.  They all use data from the first example study in the original O'Quigley paper.

```{r, subjectGraph, echo=FALSE}
data("oQuigleyPatientData")
patientSummary <- patientSummary(oQuigleyPatientData)

patientSummary %>% 
  kable(col.names=c("Dose", "Treated", "Evaluable", "No Tox", "DLT")) %>% 
  add_header_above(c(" "=1, "Number of patients"=2, "Status"=2))

subjectGraph(oQuigleyPatientData, showCohort=FALSE) + labs(title="Patient allocations and toxicity status")
```
```{r, allocationPlot, echo=FALSE}
allocationPlot(oQuigleyPatientData, doseGrid=1:6) + labs(title="Patient toxicity status by dose")

data("oQuigleyPosterior")
x <- oQuigleyPosterior %>% tidy() %>% augmentOQuigleyData() %>% filter(Parameter == "Prob")
summary <- createDoseSummary(x)
```
```{r, echo=FALSE}
patientSummary %>% 
  right_join(summary, by="Dose") %>% 
  select(Dose, Treated, Evaluable, None, Tox, Mean, Q1:Q5) %>% 
  replace(is.na(.), 0) %>%
  kable(
    digits=2,
    col.names=c("Dose", "Treated", "Evaluable", "No Tox", "Tox", "Mean", "5%", "10%", "Median", "90%", "95%")
  ) %>%
  add_header_above(c(" "=1, "Patients"=4, " "=1, "Quantiles"=5)) %>%
  add_header_above(c(" "=5, "p(Tox)"=6))
```
```{r, doseToxicityGraph, echo=FALSE}
x <- oQuigleyPosterior %>% tidy() %>% augmentOQuigleyData()
y <- x %>% filter(Parameter == "Prob")
y <- createDoseSummary(y)
doseToxicityGraph(y, targetTox=0.20, pivot=TRUE) + labs(title="Posterior dose-toxicity curve")
```
```{r, posteriorPlot, echo=FALSE}
posteriorPlot(x, recDose=2) + labs(title="Posterior densities for p(Tox | Dose)")
```
